---
title: "eda"
author: "Joseph Marcus and Hussein Al-Asadi"
date: "November 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports

```{r imports, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyr)
library(reshape2)
library(dplyr)
library(geosphere)
library(viridis)
library(gridExtra)
library(MASS)
```

## Read data

Read proccessed genotypes and meta data with lat / lon and time points

```{r read_data, warning=FALSE, message=FALSE}
# read the data
load("../data/genotypes.Rdata")
G <- genotypes
snps <- colnames(G)
geno_df <- as.data.frame(t(G))
rownames(geno_df) <- NULL
geno_df$snp <- snps
meta.df <- read.table("meta.tsv", stringsAsFactors = FALSE, header=TRUE, sep = "\t") %>% mutate(time = time+2016)
```

```{r plot_map, warning=FALSE, message=FALSE}
# plot up the map
map_plot <- ggplot(meta.df, aes(x=long, y=lat, fill=time))  + 
            scale_fill_viridis(option = "C", name = "years BP") + labs(colour = "years BP") + 
            borders("world", xlim = c(-10.00,  27.00), ylim = c(39.00, 65.00),
                    fill="#bdbdbd", colour="#bdbdbd") +
            geom_point(alpha=.6, size=3, pch=21, colour="white", position = position_jitter(width = 1.0, height = 1.0)) +
            scale_size_area() + 
            coord_map() + 
            theme_minimal() 
map_plot  + ggsave("figures/sample_locations_times.pdf", width=6, height=4)
```

## Gather data

Gather data into long format where each row snp | iid | genotype

```{r gather_data, warning=FALSE, message=FALSE}
gath_geno_df <- geno_df %>% gather(iid, genotype, -snp)
head(gath_geno_df)
```

## Merge data

Merge the lat / lon and time meta data with the genotypes

```{r merge_data, warning=FALSE, message=FALSE}
merged_gath_geno_df <- gath_geno_df %>% inner_join(meta.df, by=c("iid"))
head(merged_gath_geno_df)
```


## Filter data

Remove outlier individuals and sample genotypes from modern individuals.

```{r filter_data, warning=FALSE, message=FALSE}
# outlier individual ids
outlier_iids <- c("ATP2")

# ancient diploids
dip.ancients <- c("Stuttgart", "Loschbour")

# apply filters and read sampling
fil_merged_gath_geno_df <- merged_gath_geno_df %>%
                           filter(!(iid %in% outlier_iids)) %>% # remove outliers
                           mutate(is_modern = ifelse(time == 0, 1, 0)) %>% # add time indicator
                           mutate(sampled_genotype = ifelse((is_modern == 1 & genotype == 1) | (iid %in% dip.ancients & genotype == 1), # sample genotypes for moderns
                                                             sample(c(0, 2), size=1), 
                                                             genotype)) %>%
                           mutate(sampled_genotype = ifelse(is.na(genotype), # divide by 2 make genotypes 0, 1 variables
                                                            NA, 
                                                            sampled_genotype / 2))
```

## Calculate missingness

Calculate genotype missingness for each snp in ancients and moderns seperatly

```{r calc_miss, warning=FALSE, message=FALSE}
miss_df <- fil_merged_gath_geno_df %>% 
           group_by(snp, is_modern) %>% 
           summarise(miss_rate = sum(is.na(genotype)) / n()) 

miss_plot <- ggplot(miss_df %>% mutate(is_modern = ifelse(is_modern == 0, "Ancient", "Modern")), aes(x=miss_rate)) +     
             geom_histogram() + facet_grid(. ~ is_modern, scales = "free") + theme_bw() + xlab("Missingness rate")

miss_plot + ggsave("figures/miss_rate.pdf", height=4, width=5)

# vector of snps with high missingness rate in ancients
high_miss_df <- miss_df %>% filter(is_modern == 0, miss_rate >= .9) 
high_miss_snps <- high_miss_df$snp
```

## Create final analysis data.frame and matrix

```{r create_final_data, warning=FALSE, message=FALSE}
# filter out snps with high missigness rate in ancients in full datset
final_df <- fil_merged_gath_geno_df %>% filter(!(snp %in% high_miss_snps)) 

# create matrix for exploring ibd and ibt
final_g_mat <- final_df %>% 
               dplyr::select(snp, iid, sampled_genotype) %>% 
               spread(iid, sampled_genotype)
final_snps <- final_g_mat$snp 
final_g_mat <- as.matrix(final_g_mat %>% dplyr::select(-snp))
rownames(final_g_mat) <- final_snps
final_g_mat_t <- t(final_g_mat)
```

## Compute genetic, geographic and time distances

compute pairwise distances between all individuals

```{r compute_dist, warning=FALSE, message=FALSE}
compute_dist <- function(X, dist_type, iids=NULL){
  if(dist_type == "geo_dist"){
    D <- as.matrix(distm(X)) / 1000
  } else if(dist_type %in% c("genetic_dist", "time_dist")){
    D <- as.matrix(dist(X))  
  } else {
    stop("distance type not found")
  }
  if(dist_type %in% c("geo_dist", "time_dist")){
    colnames(D) <- iids
    rownames(D) <- iids    
  }
  diag(D) <- NA
  D[lower.tri(D)] <- NA
  df <- arrange(melt(D), value) %>% filter(!is.na(value))
  colnames(df) <- c("iid_0", "iid_1", dist_type)
  return(df)
}

# meta data.frame with outlier individuals removed
iids <- colnames(final_g_mat)

# compute distances
genetic_dist_df <- compute_dist(final_g_mat_t, "genetic_dist")
geo_dist_df <- compute_dist(as.matrix(meta.df[, c("long", "lat")]), "geo_dist", meta.df$iid)
time_dist_df <- compute_dist(as.matrix(meta.df[, c("time")]), "time_dist", meta.df$iid)

# merge distances
dist_df <- genetic_dist_df %>% 
           inner_join(geo_dist_df, by=c("iid_0", "iid_1")) %>%
           inner_join(time_dist_df, by=c("iid_0", "iid_1")) %>%
           inner_join(meta.df, by = c("iid_0" = "iid")) %>% 
           inner_join(meta.df, by = c("iid_1" = "iid"))
```

## Visualizing IBD and IBT

### Geo vs Genetic vs Time Distance

```{r geo_time_genetic_dist}
p <- ggplot(dist_df, aes(x=geo_dist, y=genetic_dist, color=time_dist)) + 
     geom_point()  + ylim(120,200) +
     scale_color_viridis() + theme_bw() + xlab("Geographic distance (km)") + ylab("Genetic distance") + labs(color = "Time distance (years)") 
p + ggsave("figures/geo_time_genetic_dist.pdf", height=4, width=6)
```

### Modern IBD

```{r mod_geo_vs_genetic_dist}
fil_dist_df <- dist_df %>% filter(time.x == 0 & time.y == 0 & genetic_dist > 90)
lm <- lm(genetic_dist ~ geo_dist, data=fil_dist_df)
r2 <- round(summary(lm)$r.squared,2)
p <- ggplot(fil_dist_df, aes(x=geo_dist, y=genetic_dist)) + geom_point(alpha=.5) +  stat_smooth(method = "lm", color="orange") + theme_bw() + xlab("Geographic distance (km)") + ylab("Genetic distance") + ggtitle(paste0("Moderns, r2 = ", r2))
p + ggsave("figures/mod_geo_vs_genetic_dist.pdf", width=6, height=4)
```


### Ancient IBD over time

```{r anc_ibd_over_time}
time_dist_grid <- seq(50, 5000, 50)
r2s <- rep(NA, length(time_dist_grid))
npoints <- rep(NA, length(time_dist_grid))
for(t in 1:length(time_dist_grid)){
  fil_dist_df_t <- dist_df %>% filter(time.x != 0 & time.y != 0, time_dist < time_dist_grid[t])
  lm_t <- lm(genetic_dist ~ geo_dist, data=fil_dist_df_t)
  r2s[t] <- summary(lm_t)$r.squared
  npoints[t] <- nrow(fil_dist_df_t)
}
r2_df <- data.frame(r2=r2s, time_dist_bin=time_dist_grid, npoints = npoints)
p1 <- ggplot(r2_df, aes(x=time_dist_bin, y=r2)) + geom_point(color="red", alpha=.7)+ geom_line() + theme_bw() +
     guides(alpha=FALSE) + xlab("Time distance < t (years)") 

p2 <- ggplot(r2_df, aes(x=time_dist_bin, y=npoints)) + geom_point(color="red", alpha=.7) + theme_bw() +
     guides(alpha=FALSE) + xlab("Time distance < t (years)") + scale_y_continuous(breaks=c(200,1000, 2000, 3000))+
  ylab("number of points in time bin")
tt = grid.arrange(p1, p2, nrow=1, ncol=2) 
ggsave("figures/anc_ibd_over_time.pdf", width=6, height=4, plot=tt)
```

### Ancient IBT over space

```{r anc_ibt_over_geo}
geo_dist_grid <- seq(30, 2500, 30)
r2s <- rep(NA, length(geo_dist_grid))
npoints <- rep(NA, length(geo_dist_grid))

for(t in 1:length(geo_dist_grid)){
  fil_dist_df_t <- dist_df %>% filter(time.x != 0 & time.y != 0, geo_dist < geo_dist_grid[t])
  lm_t <- lm(genetic_dist ~ time_dist, data=fil_dist_df_t)
  r2s[t] <- summary(lm_t)$r.squared
    npoints[t] <- nrow(fil_dist_df_t)
}
r2_df <- data.frame(r2=r2s, geo_dist_bin=geo_dist_grid, npoints = npoints)
p1 <- ggplot(r2_df, aes(x=geo_dist_bin, y=r2)) + geom_point(color="red", alpha=.7)+ geom_line() + theme_bw() + guides(alpha=FALSE) + xlab("Geographic distance < d (km)")
p2 <- ggplot(r2_df, aes(x=geo_dist_bin, y=npoints)) + geom_point(color="red", alpha=.7) + theme_bw() +
     guides(alpha=FALSE) + scale_y_continuous(breaks=c(450,1000, 2000, 3000)) + xlab("Geographic distance < d (km)") + ylab("number of points in geo. bin")
tt = grid.arrange(p1, p2, nrow=1, ncol=2)
ggsave("figures/anc_ibd_over_geo.pdf", width=6, height=4, plot=tt)
```

### Ancient IBT for different countries

```{r}

countries <- c("Germany", "Hungary", "Spain", "Sweden")
plots <- list()
r2s <- rep(NA, length(countries))

for (cntry in 1:length(countries)){
  iids <- meta.df$iid[which(meta.df$country==countries[cntry] & meta.df$source != "GuntherPNAS2015")]
  cntry_dist_df <- dist_df %>% filter(iid_0 %in% iids) %>%  filter(iid_1 %in% iids)
  p <- ggplot(cntry_dist_df, aes(x=time_dist, y=genetic_dist)) + geom_point(alpha=.5) +  stat_smooth(method = "lm", color="orange", se=FALSE) + theme_bw() + xlab("time distance (years)") + ylab("Genetic distance")
  lm_cntry <- lm(genetic_dist ~ time_dist, data = cntry_dist_df)
  r2s[cntry] <- round(summary(lm_cntry)$r.squared,3)
  p <- p + ggtitle(paste0(countries[cntry], ", r^2 = ", r2s[cntry]))
  plots[[cntry]] = p
}

plot.to.save = grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]], nrow=2, ncol=2)
ggsave("figures/ibt_bycountry.pdf", height=4, width=6, plot = plot.to.save)
```


```{r}
genetic.dist.m <- as.matrix(dist(final_g_mat_t))
mds = isoMDS(genetic.dist.m)
plot(mds$points, type = "n", pch = 20, cex = 3, col = adjustcolor("black", alpha = 0.3), xlab = "X", ylab = "Y") 
my.df <- data.frame(iid = row.names(mds$points), x = mds$points[,1], y = mds$points[,2])
mds.df = left_join(my.df, meta.df)
gg <- ggplot(mds.df, aes(x=x, y=y, fill=time))  + 
            scale_fill_viridis(option = "C") + labs(color = "years BP", title = "MDS", x = "coordinate 1", y = "coordinate 2")  +
            geom_point(alpha=.6, size=3, pch=21, colour="white") +
            theme_minimal() 
gg + ggsave("figures/mds.pdf", width=6, height=4)
```
