---
title: "eda"
author: "Joseph Marcus and Hussein Al-Asadi"
date: "November 15, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports

```{r imports, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyr)
library(reshape2)
library(dplyr)
library(geosphere)
library(viridis)
```

## Read data

Read proccessed genotypes and meta data with lat / lon and time points

```{r read_data, warning=FALSE, message=FALSE}
# read the data
load("../data/genotypes.Rdata")
G <- genotypes
snps <- colnames(G)
geno_df <- as.data.frame(t(G))
rownames(geno_df) <- NULL
geno_df$snp <- snps
meta_df <- read.table("../data/meta.tsv", stringsAsFactors = FALSE, header=TRUE) %>% mutate(time=time+2016)

# plot up the map
map_plot <- ggplot(meta_df, aes(x=long, y=lat, fill=time)) + 
            scale_fill_viridis(option = "C") + 
            borders("world", xlim = c(-10.00,  27.00), ylim = c(39.00, 65.00),
                    fill="#bdbdbd", colour="#bdbdbd") + 
            geom_point(alpha=.6, size=3, pch=21, colour="white", position = position_jitter(width = 4, height = 4)) +
            scale_size_area() + 
            coord_map() + 
            theme_minimal()
map_plot + ggsave("figures/sample_locations_times.pdf", width=6, height=4)
```

## Gather data

Gather data into long format where each row snp | iid | genotype

```{r gather_data, warning=FALSE, message=FALSE}
gath_geno_df <- geno_df %>% gather(iid, genotype, -snp)
head(gath_geno_df)
```

## Merge data

Merge the lat / lon and time meta data with the genotypes

```{r merge_data, warning=FALSE, message=FALSE}
merged_gath_geno_df <- gath_geno_df %>% inner_join(meta_df, by=c("iid"))
head(merged_gath_geno_df)
```


## Filter data

Remove outlier individuals and sample genotypes from modern individuals.

```{r filter_data, warning=FALSE, message=FALSE}
# outlier individual ids
outlier_iids <- c("ATP2", "Stuttgart", "Loschbour")

# apply filters and read sampling
fil_merged_gath_geno_df <- merged_gath_geno_df %>%
                           filter(!(iid %in% outlier_iids)) %>% # remove outliers
                           mutate(is_modern = ifelse(time == 0, 1, 0)) %>% # add time indicator
                           mutate(sampled_genotype = ifelse((is_modern == 1 & genotype == 1), # sample genotypes for moderns
                                                             sample(c(0, 2), size=1), 
                                                             genotype)) %>%
                           mutate(sampled_genotype = ifelse(is.na(genotype), # divide by 2 make genotypes 0, 1 variables
                                                            NA, 
                                                            sampled_genotype / 2))
```

## Calculate missingness

Calculate genotype missingness for each snp in ancients and moderns seperatly

```{r calc_miss, warning=FALSE, message=FALSE}
miss_df <- fil_merged_gath_geno_df %>% 
           group_by(snp, is_modern) %>% 
           summarise(miss_rate = sum(is.na(genotype)) / n()) 

miss_plot <- ggplot(miss_df %>% mutate(is_modern = ifelse(is_modern == 0, "Ancient", "Modern")), aes(x=miss_rate)) +     
             geom_histogram() + facet_grid(. ~ is_modern, scales = "free") + theme_bw() + xlab("Missingness rate")

miss_plot + ggsave("figures/miss_rate.pdf", height=4, width=5)

# vector of snps with high missingness rate in ancients
high_miss_df <- miss_df %>% filter(is_modern == 0, miss_rate >= .6) 
high_miss_snps <- high_miss_df$snp
```

## Create final analysis data.frame and matrix

```{r create_final_data, warning=FALSE, message=FALSE}
# filter out snps with high missigness rate in ancients in full datset
final_df <- fil_merged_gath_geno_df %>% filter(!(snp %in% high_miss_snps)) 

# create matrix for exploring ibd and ibt
final_g_mat <- final_df %>% 
               select(snp, iid, sampled_genotype) %>% 
               spread(iid, sampled_genotype)
final_snps <- final_g_mat$snp 
final_g_mat <- as.matrix(final_g_mat %>% select(-snp))
rownames(final_g_mat) <- final_snps
final_g_mat_t <- t(final_g_mat)
```

## Compute genetic, geographic and time distances

compute pairwise distances between all individuals

```{r compute_dist, warning=FALSE, message=FALSE}
compute_dist <- function(X, dist_type, iids=NULL){
  if(dist_type == "geo_dist"){
    D <- as.matrix(distm(X)) / 1000
  } else if(dist_type %in% c("genetic_dist", "time_dist")){
    D <- as.matrix(dist(X))  
  } else {
    stop("distance type not found")
  }
  if(dist_type %in% c("geo_dist", "time_dist")){
    colnames(D) <- iids
    rownames(D) <- iids    
  }
  diag(D) <- NA
  D[lower.tri(D)] <- NA
  df <- arrange(melt(D), value) %>% filter(!is.na(value))
  colnames(df) <- c("iid_0", "iid_1", dist_type)
  return(df)
}

# meta data.frame with outlier individuals removed
iids <- colnames(final_g_mat)

# compute distances
genetic_dist_df <- compute_dist(final_g_mat_t, "genetic_dist")
geo_dist_df <- compute_dist(as.matrix(meta_df[, c("long", "lat")]), "geo_dist", meta_df$iid)
time_dist_df <- compute_dist(as.matrix(meta_df[, c("time")]), "time_dist", meta_df$iid)

# merge distances
dist_df <- genetic_dist_df %>% 
           inner_join(geo_dist_df, by=c("iid_0", "iid_1")) %>%
           inner_join(time_dist_df, by=c("iid_0", "iid_1")) %>%
           inner_join(meta_df, by = c("iid_0" = "iid")) %>% 
           inner_join(meta_df, by = c("iid_1" = "iid"))
```

## Visualizing IBD and IBT

### Geo vs Genetic vs Time Distance

```{r geo_time_genetic_dist}
p <- ggplot(dist_df, aes(x=geo_dist, y=genetic_dist, color=time_dist)) + 
     geom_point() + 
     scale_color_viridis() + theme_bw() + xlab("Geographic distance (km)") + ylab("Genetic distance") + labs(color = "Time distance (years)") 
p + ggsave("figures/geo_time_genetic_dist.pdf", height=4, width=6)
```

```{r geo_time_dist_dens}
p <- ggplot(dist_df, aes(x=time_dist)) + geom_density(fill="red", alpha=.5) + theme_bw() + xlab("Time distance (years)") 
p + ggsave("figures/time_dist_dens.pdf", width=6, height=4)

p <- ggplot(dist_df, aes(x=geo_dist)) + geom_density(fill="orange", alpha=.5) + theme_bw() + xlab("Geographic distance (km)") 
p + ggsave("figures/geo_dist_dens.pdf", width=6, height=4)
```

### Modern IBD

```{r mod_geo_vs_genetic_dist}
fil_dist_df <- dist_df %>% filter(time.x == 0 & time.y == 0 & genetic_dist > 90)
p <- ggplot(fil_dist_df, aes(x=geo_dist, y=genetic_dist)) + geom_point(alpha=.5) +  stat_smooth(method = "lm", color="orange") + theme_bw() + xlab("Geographic distance (km)") + ylab("Genetic distance")
p + ggsave("figures/mod_geo_vs_genetic_dist.pdf", width=6, height=4)
```


### Ancient IBD over time

```{r anc_ibd_over_time}
time_dist_grid <- seq(50, 8000, 50)
r2s <- rep(NA, length(time_dist_grid))
for(t in 1:length(time_dist_grid)){
  fil_dist_df_t <- dist_df %>% filter(time.x != 0 & time.y != 0, time_dist < time_dist_grid[t])
  lm_t <- lm(genetic_dist ~ geo_dist, data=fil_dist_df_t)
  r2s[t] <- summary(lm_t)$r.squared
}
r2_df <- data.frame(r2=r2s, time_dist_bin=time_dist_grid)
p <- ggplot(r2_df, aes(x=time_dist_bin, y=r2)) + geom_point(color="red", alpha=.7)+ geom_line() + theme_bw() +
     guides(alpha=FALSE) + xlab("Time distance < t (years)") 
p + ggsave("figures/anc_ibd_over_time.pdf", width=6, height=4)
```

### Ancient IBT over space

```{r anc_ibt_over_geo}
geo_dist_grid <- seq(25, 4000, 25)
r2s <- rep(NA, length(geo_dist_grid))
for(t in 1:length(geo_dist_grid)){
  fil_dist_df_t <- dist_df %>% filter(time.x != 0 & time.y != 0, geo_dist < geo_dist_grid[t])
  lm_t <- lm(genetic_dist ~ time_dist, data=fil_dist_df_t)
  r2s[t] <- summary(lm_t)$r.squared
}
r2_df <- data.frame(r2=r2s, geo_dist_bin=geo_dist_grid)
p <- ggplot(r2_df, aes(x=geo_dist_bin, y=r2)) + geom_point(color="orange", alpha=.7)+ geom_line() + theme_bw() +
     guides(alpha=FALSE) + xlab("Geographic distance < d (km)")
p + ggsave("figures/anc_ibt_over_geo.pdf", width=6, height=4)
```

### PCA of Ancients

```{r}
replace_missing = function(x){
  x[is.na(x)] <- mean(x, na.rm=TRUE)
  return(x)
}

anc_final_df <- final_df %>% filter(is_modern == 0)
anc_g_final_df <- anc_final_df %>% select(snp, iid, sampled_genotype) %>% spread(iid, sampled_genotype) 
snps <- anc_g_final_df$snp
anc_g_mat <- as.matrix(anc_g_final_df %>% select(-snp))
rownames(anc_g_mat) <- snps
anc_g_mat <- t(apply(anc_g_mat, 1, replace_missing))
sds <- apply(anc_g_mat, 1, sd)
anc_g_mat_fil <- anc_g_mat[-which(sds == 0),]
pca <- prcomp(t(anc_g_mat_fil), center = TRUE, scale. = TRUE)
pca_df <- as.data.frame(pca$x)
pca_df$iid <- rownames(pca_df)
pca_df <- pca_df %>% inner_join(meta_df, by=c("iid"))
vars <- (pca$sdev)^2 
pve <- vars / sum(vars)
p <- ggplot(pca_df, aes(x=PC1, y=PC2, color=time)) + geom_point() + 
     scale_color_viridis() + xlab(paste0("PC1 (variance explained=", sprintf("%.03f%%", pve[1]*100), ")")) +
     ylab(paste0("PC2 (variance explained=", sprintf("%.03f%%", pve[2]*100), ")")) + theme_bw()
p + ggsave("figures/anc_pca.pdf", width=6, height=4)
```
